Machine Learning Tools for Open Cluster Characterization with Gaia DR2 Data

### 2021-02-09 thesis/thesis.tex

\usepackage{apacite}
\usepackage{caption}
\usepackage{csquotes}
\usepackage{eso-pic}
\usepackage{float}
\usepackage{graphicx}
\usepackage{minted}
\usepackage{picture}
\usepackage{subcaption}
\usepackage{tablefootnote}
\usepackage[nottoc]{tocbibind}

@@

This research has made use of the VizieR catalogue access tool, CDS, Strasbourg, France (DOI : 10.26093/cds/vizier).
The original description of the VizieR service was published in 2000, \cite[A\&AS 143, 23]{ochsenbein2000vizier}

@@

Stellar open clusters (OCs) \cite{janes1982open} are groups of stars gravitationally bounded originated from a single molecular gas cloud.
They share the same chemical composition and age. Moreover, they have similar relative positions and proper motion.
Those astronomical objects are relevant to understand the spiral structure, the dynamics and the chemical evolution of our galaxy.

Although most stars in the Milky Way are presented isolated, it is considered that most of them (or even all)
are formed in clustered environments and spend a period of time gravitationally bounded with their siblings embedded
in their original molecular cloud \cite{clarke2000theformationofstellarclusters} \cite{portegies2010young}.
The evolution of those systems tends to sparse them in a few million years by interacting gravitationally with other systems.
Galactic tidal forces and mechanisms that involve gas loss driven by stellar feedback are other causes of disruption
\cite{brinkmann2017bound}.
Nevertheless, a small fraction of those systems will survive in the initial state and persist bounded in larger timescales.

@@

The study of OCs has been pushed forward thanks to the huge and precise dataset from the Gaia mission
\cite{collaboration2016description} Gaia DR2 \cite{gaia2018gaia}, available since 2018.
This dataset has helped to review already known open clusters and to find new ones.

@@

We will avoid this issue by looking at $E(b-r) / G_{mag}$ diagrams that show their connection with the isochrone curves.
These curves are a tool used to determine how stars evolve in time \cite{bressan2012parsec}.
The isochrones are derived from theoretical models.
These models are mainly based on metallicity and mass/brightness ratio of stars.
They also have a direct relation with that presented by the Hertzsprung–Russell (H-R) diagram of the stars belonging to the cluster.
Examples of these curves are shown in Figure \ref{fig:examples_of_isochrones}.

@@

Since these stars were born close in time, they should have a sharp and well-defined profile,
with low scattering through the main sequence in the H-R diagram.
Examples for this kind of diagrams are presented in Figure \ref{fig:examples_of_hr_diagrams}.

@@

Although the members of a cluster appear in the observational visual field as an overdensity in positions,
as shown in Figure \ref{fig:pos_ngc_2682},
these coordinates are not useful to separate those stars that belong to the cluster from the other that do not.
However, if we look for overdensities in the proper motion configuration spaces, it is possible, at least at first instance,
to assume a possible membership cut (see Figure \ref{fig:pm_ngc_2682}.

@@

In many other cases things are not so simple,
and the characterization process can turn very difficult and tedious using tools such as the available ones at the Virtual Observatory (VO).
Furthermore, it frequently requires a parametrization based on prior knowledge regarding the studied region.
The aim of this work is to test an unsupervised and non-parametrized model based on Machine Learning (ML) tools.

@@

The primary aim of this thesis is to \emph{build an unsupervised clustering model for open cluster characterization}.
The model will be \emph{non-supervised and non-parameterized} to fit a wide range of clusters without
the need for fine-tuning a high number of hyperparameters.

@@

An initial approach for finding OCs is the search for overdensities along the galactic disk.
In general, this is a good starting point but, although it seems simple,
it presents a fundamental problem already discussed.
The near field around the OC is filled with two types of distinct star populations:
those who belong to the OC (tens or hundreds to a few thousand)
and a background made up of thousands or millions of stars that are not.
Finding out which stars belong to the first group is the problem faced in this work.
This selection is crucial to properly characterize the fundamental properties of the cluster
(dynamics, total mass, age, chemical composition, etc.).

@@

Figure \ref{fig:pm_melotte_22} shows the proper motion distribution in right ascension and declination.
It is easy to see that a subgroup can be located at [20, -45].
Figure \ref{fig:pm_vec_melotte_22} shows the parallax distribution and confirms an overdensity at
$\approx 7.3 mas$ corresponding to Mellote 22 OC (See also Figure \ref{fig:melotte_22_parallax_zoom}).

However, as shown in Figures \ref{fig:pm_ngc_6494} and \ref{fig:parallax_ngc_6494},
in general it is not as easy and becomes necessary to consider other parameters such as distances,
or even metallicity and age (derived from isochrone curves).
Sometimes even, photometric data may be required for the stars within the studied field.

@@

In this work we make use of Gaia DR2 since DR3 has not been released in time for us to include it.
Gaia DR2 is a multidimensional dataset obtained by ESA's Gaia mission (located at L2, 1.5 million kilometers from Earth) and
operational since 2014. The catalogue has high precision and accuracy astrometric data for more than 1.7 billion stellar sources,
and magnitudes in three photometric filters (G, BP and RP) for more than 1,300 million sources.
80\% of Gaia DR2 sources are weaker than $G_{mag} \approx 18$.
For magnitudes $G \approx 21$ nominal uncertainties reach $2 mas$ for parallaxes and $5 mas \cdot years^{-1}$ in proper motions.
But for closer sources, (at the bright end, $G < 14$), the precision is of the order of $0.02 mas$
and $0.05 mas \cdot year^{-1}$ respectively \cite{cantat2018gaia}.
In general, open clusters, which are the main subject of this work, are at optimal data thresholds.

In this context, and taking advantage of the opportunity provided by Gaia DR2,
new approaches are being developed to improve current detection methods and automating OC characterization as much as possible.
This is the case of tools such as Clusterix 2.0 \cite{balaguer2020clusterix}, TOPCAT \cite{taylor2005topcat}
and VOSA \cite{bayo2008vosa}, (all of them developed by the Virtual Observatory).
These tools work together and offer good results,
although it is not possible to use them in an unsupervised process and they require different parameterizations for each study case.

@@

Another tool is Clusterix 2.0, which is an interactive web-based tool \cite{balaguer2020clusterix}.
It takes the proper motion diagram without making any prior assumption about the membership of the candidate star
and determines empirically the frequency functions.

@@

\begin{equation}
  K(a, b) = \frac{1}{2 \pi h^{2}} \exp{ \left[ - \frac{1}{2}\frac{\left( a - a_{i} \right)^{2} + \left( b - b_{i} \right)^{2}}{ h^{2}} \right]}
\end{equation}

$\left( a, b \right)$ are referred to the proper motion configuration space,
while a point located in the center of the cell $\left( i, j \right)$ provides the maximum
contribution for calculating the local density.
Number $h$ is called the \emph{smoothing} parameter and it is measured in the same units as proper motion.

Various clustering algorithms have been studied in the past with the purpose of finding a valid non-parameterized method.
UPMASK \cite{krone2014upmask} was developed to only use photometry and positions,
and then it was adapted to the Gaia data \cite{cantat2018gaia} based only on proper motion and parallax.
A similar approach has been carried out using DBSCAN and machine learning algorithms to discover new clusters.

@@

\begin{displayquote}
The non-field population does not occupy the entire workspace, but is spatially concentrated,
which makes it possible to distinguish two regions in the workspace:
the only field region (label 'f'), dominated by star fields,
and the cluster + field region (label 'c+f'), which includes both star fields and not star fields.
\cite{balaguer2020clusterix}
\end{displayquote}

This fact implies defining three areas or regions with different radius.
The first 'c+f' corresponds to the one in which the cluster members are
presumed to be contained together with other star fields that are not part of the cluster.
The second region is the broadest and assumes that it only contains stars
in an extended visual field without components of the cluster.
The third region is the intermediate one and is out of analysis (void area),
since it would correspond to a possible transition zone between the other two.

The right choice of these radii, even having a previous estimation for the 'c+f' region,
highly affects the execution of the algorithm and, in general, requires a considerable wide field 'f'.
There is no rule of thumb that defines relative proportions of these areas.

Finally, when an acceptable result is obtained,
what Clusterix provides is a probability field associated to the set of stars under study.
The recovered dataset contains all fields from Gaia database
and an extra column with the probability value of each object to belong to the open cluster.
This new dataset can be exported to TOPCAT for further optimization and final processing.
Figure \ref{fig:clusterix_probability} shows an example of the probability distribution
in the proper motion space for NGC 2682.

@@

Finally, it is necessary to refer to the work developed on \cite{castro2020hunting}.
It presents a method based on machine learning techniques to make a systematic search
for overdensities in the astrometric space of the galactic disk
and a later identification of OC using photometric information, also from Gaia DR2.

The method includes two phases: the first one uses an unsupervised clustering algorithm, DBSCAN,
to search for overdensities $(l, b \pi, \mu_{\alpha} *, \mu_{\delta})$,
and then applies a deep learning Artificial Neural Network (ANN),
previously trained with magnitude diagrams,
to identify isochrone patterns within the detected overdensities and thus proceed to confirm them as OC.

@@

The result of this work is the recognition of 582 new open clusters
distributed along the galactic disk for a galactic declination $b < 20^{\circ}$,
which has meant an increase of 45\% over previously ones.

Our aim in this work, linked to computational limitations compared to MareNostrum,
is not to do a blind search for clusters, but to obtain a method,
within the machine learning environment,
that allows the characterization of open clusters through unsupervised and non-parametrized procedures in a single step.
This method should not require further refinements with other tools.

@@

There are several clustering algorithms: K-Means, Mean-Shift Clustering, DBSCAN \cite{ester1996density} among other.
While each one behaves better according to the distribution of the objects to clusterize,
we have chosen K-Means by its simplicity and good results.

K-Means requires a single parameter, the number of clusters to build (N).
The algorithm looks for N center points which are vectors of the same dimension as the number of selected features.
K-Means starts by making an initial groups configuration and then,
it reassigns objects to other groups iteratively by minimizing the distance among points
inside the new group and maximizing the distance with the centers of the other groups.

@@

This algorithm works well and gives good results at first approach.
However, we need to set a large number of clusters to find the open cluster we are looking for.
This fact complicates the identification of the OC and many times the found cluster contains to many outliers.
For that reason, we have search for a K-Means refinement based on an artificial neural network.

The \emph{Unsupervised Deep Embedding for Clustering Analysis} model
or \emph{DEC} \cite{xie2016unsupervised} takes K-Means as its starting point,
but then, it trains an autoencoder to reduce the feature space
and pass this transformed data through a Clustering Layer which refines the previous selection.

The mechanism of this model is explained in Section \ref{sec:deep_embedding_clustering}
and its implementation is available in \verb|cdalvaro.ml.dec| package.

@@

All code has been developed with the Python programming language \cite{Python3}.
Other auxiliary tools like \emph{Docker} \cite{merkel2014docker},
\emph{PostgreSQL} \cite{postgresql}, \emph{Jupyter Notebooks} \cite{Kluyver2016jupyter},
and frameworks such as \emph{Astropy} \cite{astropy:2013} \cite{astropy:2018},
\emph{Scikit-Learn} \cite{scikit-learn}, \emph{Seaborn} \cite{michael_waskom_2017_883859},
\emph{SQLAlchemy} \cite{sqlalchemy} and \emph{Keras} \cite{chollet2015keras} have been used too.

@@

For the sake of simplicity, we are taking \emph{Melotte 22 (Messier 45)} \cite{elsanhoury2019ppmxl} to illustrate the whole method.
This cluster is well studied and its stars are not too mixed with other stars that do not belong to the OC
although they are contained in the same observation field, so images look clear.

Figure \ref{fig:raw_pm_melotte_22} shows \emph{proper motions in right ascension and declination}
for a sample of the downloaded dataset for Melotte 22.
At first sight, two main clusters can be distinguished, one of them centered nearly at (0, 0)
and the second one with center at (20, -45). This second cluster is the one we are looking for.

@@

Figure \ref{fig:raw_parallax_hr_diagram_mellote_22} shows the parallax histogram and the H-R diagram for this cluster, respectively.

The figure on the left shows a resonance at $\approx 7.3mas$ belonging the OC.
While the figure on the right would help us to look for isochrone curves and so,
to determine the age of these stars.

@@

Before being able to develop and test our clustering model,
the first step is to download the required data from Gaia repository
and store it in a custom database for later access.

Due to the large amount of data available at Gaia, a complete download is not viable neither useful.
In order to reduce the size of the dataset to be downloaded,
the OpenClust catalogue \cite{dias2002new} (Figure \ref{fig:OpenClustComplete})
has been used to restrict the sky regions to be explored.

@@

As shown in Figure \ref{fig:OpenClustSelection},
these constrains give us a smaller dataset but it is still a good representation of all clusters in the Milky Way
since they are equally distributed around the galaxy disk.

@@

The first one is a container built from a \verb|python:3.8| image that contains the \verb|cdalvaro| package
and the \verb|downloader.py| script.
This script is prepared to load the OpenClust catalogue, connect to the Gaia DR2 database,
download all stars contained inside each cluster (with a radius multiplier to increase the area to be downloaded)
and save the downloaded data inside our custom database hosted in the second container.

@@

These containers are orchestrated with a \verb|docker-compose.yml| file which automatically launches one instance of each container.
Figure \ref{fig:services_diagram} shows a diagram describing the services architecture for this work.

@@

The second table is partitioned by region,
storing all stars related to one region inside an independent table for optimized access performance.
Partition tables are located inside \verb|gaiadr2| schema. See Figure \ref{fig:customdb_diagram}.

@@

Table \ref{tab:important_fields} shows a brief description of the fields used in this work.

@@

As mentioned before, we want our model to characterize open clusters by looking at their dynamic properties.

Also, we want to maintain as simple as possible our clustering model in order to save computing resources.

@@

Parallax is another important feature. It lets us know how far stars are from us.
In addition, since all stars within an open cluster were born from the same dust cloud,
they must all have similar parallax.

@@

Figure \ref{fig:features_melotte_22} shows a pairwise relationship among some features available
in the dataset for Melotte 22 data.
The diagonal of the grid shows the marginal distribution of the data in each column.
This plot can be useful to find correlations between pairs of variables or
to identify resonances in certain variables such as those in the parallax feature.

@@

\begin{itemize}
  \item \emph{Proper motion in right ascension} (corrected by parallax): $\mu_{\alpha}$
  \item \emph{Proper motion in declination} (corrected by parallax): $\mu_{\delta}$
  \item \emph{Parallax}: $\varpi$
  \item \emph{Proper motion modulus}: $\left\| \vec{\mu} \right\|$
\end{itemize}

This feature selection has been refined by iterating over the K-Means clustering process
with a fixed number of clusters and varying the feature selection in each case.
The final selection is the one with best \emph{silhouette score} \cite{rousseeuw1987silhouettes}.

@@

Once we have found the set of features that best describes our problem,
we can begin searching for the OC within the study region.

@@

As shown in Figure \ref{fig:kmeans_comparisons_melotte_22},
larger values for the number of clusters allow us to isolate more accurately the resonance in parallax at $\approx 7.3mas$.
However, we have the disadvantage that more groups are formed.

@@

K-Means does a good job making an initial clustering,
as shown in Figure \ref{fig:kmeans_melotte_22}.
However, too many clusters arise from this characterization
and the OC is still polluted with stars that do not belong to it.
Moreover, we would like to reduce the amount of clusters too.

@@

If we take a look into the H-R diagram (Figure \ref{fig:kmeans_hr_diagram_melotte_22}),
we can identify the isochrone curve for Melotte 22 cluster.
This curve has a good shape, but again, it contains outsider stars.
Therefore, we would like to find a better model that improves this initial characterization
by reducing the amount of clusters and also that removes outsiders from the OC.

@@

The model is composed by a \emph{deep autoencoder}, a \emph{deep encoder} and a \emph{clustering layer}.

Encoders are used to transform the input data into a latent space
using a non-linear mapping function $f_{\theta} : X \rightarrow Z$.

Although, as explained in section \ref{sec:feature_selection},
the number of features we are managing is not too large,
this latent space helps us reduce the number of features
and avoids the \emph{"curse of dimensionality"} \cite{bellman1961curse}.

These encoders are pretrained before fitting the model to generate predictions.
Then, the encoder is used with the aim of transforming input data to the latent space $Z$.
Once the data has been transformed, a K-Means clusterer is used in order to make an initial clustering.

With that initial configuration,
the model iterates alternating between computing an auxiliary target distribution (Soft Assignment)
and minimizing the Kullback-Leiber (KL) divergence \cite{kullback1951information} to it.
This unsupervised algorithm allows us to improve the clustering.

@@

In the soft assignment stage,
the \emph{Student's t-distribution} is used as a kernel to measure the similarity
between the embedded points and the cluster centroid.
While in the KL divergence minimization the algorithm iteratively refines clusters by learning
from their high confidence assignments with the help of an auxiliary target distribution.
The model is trained by matching the soft assignment to the target distribution.
The choice of this target distribution is crucial for DEC's performance.
In this work we have taken the target distribution from DEC's original paper \cite{xie2016unsupervised},
which is defined in Equation \ref{eq:student_tdistribution}.

Figure \ref{fig:dec_model_setup} shows the layer setup of our DEC model.
It is simpler than the one tested on the original paper \cite{xie2016unsupervised},
since the number of selected features in our work is smaller than in the original one.
Therefore, using the same configuration would result in a model so
powerful that would incur in overfitting issues unable to make right predictions.

@@

As shown in Figure \ref{fig:dec_melotte_22},
the DEC model is able to move stars from one group to another,
removing some clusters and improving the OC characterization.
This is exactly what we are looking for.

@@

The H-R diagram has been improved as well,
and the isochrone curve of the open cluster is sharper than the one obtained with the K-Means algorithm
(Figure \ref{fig:kmeans_hr_diagram_melotte_22}).

Figure \ref{fig:hr_diagram_comparison} shows a comparison between the H-R diagram obtained
by DEC clustering and the one obtained using the Clusterix + TOPCAT method.

@@

Finally, we can refine this selection by filtering those stars which are bellow and above
the 0.10 and 0.90 quantiles for each group, respectively.
That way we remove the most doubtful values from the selection.
Figure \ref{fig:melotte_22_filtered} shows the result after filtering DEC clustering selection.

>>>>

\usepackage{apacite}
\usepackage{caption}
\usepackage{csquotes}
\usepackage{eso-pic}
\usepackage{float}
\usepackage{graphicx}
\usepackage{minted}
\usemintedstyle{friendly}
\usepackage{picture}
\usepackage{subcaption}
\usepackage{tablefootnote}
\usepackage[nottoc]{tocbibind}

@@

This research has made use of the VizieR catalogue access tool, CDS, Strasbourg, France (DOI : 10.26093/cds/vizier).
The original description of the VizieR service was published in 2000~\cite[A\&AS 143, 23]{ochsenbein2000vizier}

@@

Stellar open clusters (OCs)~\cite{janes1982open} are groups of stars gravitationally bounded originated from a single molecular gas cloud.
They share the same chemical composition and age. Moreover, they have similar relative positions and proper motion.
Those astronomical objects are relevant to understand the spiral structure, the dynamics and the chemical evolution of our galaxy.

Although most stars in the Milky Way are presented isolated, it is considered that most of them (or even all)
are formed in clustered environments and spend a period of time gravitationally bounded with their siblings embedded
in their original molecular cloud~\cite{clarke2000theformationofstellarclusters}~\cite{portegies2010young}.
The evolution of those systems tends to sparse them in a few million years by interacting gravitationally with other systems.
Galactic tidal forces and mechanisms that involve gas loss driven by stellar feedback are other causes of disruption
\cite{brinkmann2017bound}.
Nevertheless, a small fraction of those systems will survive in the initial state and persist bounded in larger timescales.

@@

The study of OCs has been pushed forward thanks to the huge and precise dataset from the Gaia mission
\cite{collaboration2016description} Gaia DR2~\cite{gaia2018gaia}, available since 2018.
This dataset has helped to review already known open clusters and to find new ones.

@@

We will avoid this issue by looking at $E(b-r) / G_{mag}$ diagrams that show their connection with the isochrone curves.
These curves are a tool used to determine how stars evolve in time~\cite{bressan2012parsec}.
The isochrones are derived from theoretical models.
These models are mainly based on metallicity and mass/brightness ratio of stars.
They also have a direct relation with that presented by the Hertzsprung–Russell (H-R) diagram of the stars belonging to the cluster.
Examples of these curves are shown in Figure~\ref{fig:examples_of_isochrones}.

@@

Since these stars were born close in time, they should have a sharp and well-defined profile,
with low scattering through the main sequence in the H-R diagram.
Examples for this kind of diagrams are presented in Figure~\ref{fig:examples_of_hr_diagrams}.

@@

Although the members of a cluster appear in the observational visual field as an overdensity in positions,
as shown in Figure~\ref{fig:pos_ngc_2682},
these coordinates are not useful to separate those stars that belong to the cluster from the other that do not.
However, if we look for overdensities in the proper motion configuration spaces, it is possible, at least at first instance,
to assume a possible membership cut (see Figure~\ref{fig:pm_ngc_2682}.

@@

In many other cases things are not so simple,
and the characterization process can turn very difficult and tedious using tools such as the available ones at the Virtual Observatory (VO).
Furthermore, it frequently requires a parametrization based on prior knowledge regarding the studied region.
The aim of this work is to test an unsupervised and non-parameterized model based on Machine Learning (ML) tools.

@@

The primary aim of this thesis is to \emph{build an unsupervised clustering model for open cluster characterization}.
The model must be \emph{non-supervised and non-parameterized} to fit a wide range of clusters without
the need for fine-tuning a high number of hyperparameters.

@@

An initial approach for finding OCs is the search for overdensities along the galactic disk.
In general, this is a good starting point but, although it seems simple,
it presents a fundamental problem already discussed.
The near field around the OC is filled with two types of distinct star populations:
those who belong to the OC (tens or hundreds to a few thousand)
and a background made up of thousands or millions of stars that are not.
Finding out which stars belong to the first group is the problem faced in this work.
This selection is crucial to properly characterize the fundamental properties of the cluster
(dynamics, total mass, age, chemical composition, among other).

@@

Figure~\ref{fig:pm_melotte_22} shows the proper motion distribution in right ascension and declination.
It is easy to see that a subgroup can be located at [20, -45].
Figure~\ref{fig:pm_vec_melotte_22} shows the parallax distribution and confirms an overdensity at
\(\approx 7.3 mas\) corresponding to Mellote 22 OC (See also Figure~\ref{fig:melotte_22_parallax_zoom}).

However, as shown in Figures~\ref{fig:pm_ngc_6494} and~\ref{fig:parallax_ngc_6494},
in general it is not as easy and becomes necessary to consider other parameters such as distances,
or even metallicity and age (derived from isochrone curves).
Sometimes even, photometric data may be required for the stars within the studied field.

@@

In this work we make use of Gaia DR2 since DR3 has not been released in time for us to include it.
Gaia DR2 is a multidimensional dataset obtained by ESA's Gaia mission (located at L2, 1.5 million kilometers from Earth) and
operational since 2014. The catalogue has high precision and accuracy astrometric data for more than 1.7 billion stellar sources,
and magnitudes in three photometric filters (G, BP and RP) for more than 1,300 million sources.
80\% of Gaia DR2 sources are weaker than \(G_{mag} \approx 18\).
For magnitudes \(G \approx 21\) nominal uncertainties reach \(2 mas\) for parallaxes and \(5 mas \cdot years^{-1}\) in proper motions.
But for closer sources, (at the bright end, \(G < 14\)), the precision is of the order of \(0.02 mas\)
and \(0.05 mas \cdot year^{-1}\) respectively~\cite{cantat2018gaia}.
In general, open clusters, which are the main subject of this work, are at optimal data thresholds.

In this context, and taking advantage of the opportunity provided by Gaia DR2,
new approaches are being developed to improve current detection methods and automating OC characterization as much as possible.
This is the case of tools such as Clusterix 2.0~\cite{balaguer2020clusterix}, TOPCAT~\cite{taylor2005topcat}
and VOSA~\cite{bayo2008vosa}, (all of them developed by the Virtual Observatory).
These tools work together and offer good results,
although it is not possible to use them in an unsupervised process and they require different parameterizations for each study case.

@@

Another tool is Clusterix 2.0, which is an interactive web-based tool~\cite{balaguer2020clusterix}.
It takes the proper motion diagram without making any prior assumption about the membership of the candidate star
and determines empirically the frequency functions.

@@

\begin{equation}
  K(a, b) = \frac{1}{2 \pi h^{2}} \exp{ \left[ - \frac{1}{2}\frac{\left( a - a_{i} \right)^{2} + \left( b - b_{j} \right)^{2}}{ h^{2}} \right]}
\end{equation}

\(\left( a, b \right)\) are referred to the proper motion configuration space,
while a point located in the center of the cell \(\left( i, j \right)\) provides the maximum
contribution for calculating the local density.
\(h\) is called the \emph{smoothing} parameter and it is measured in the same units as proper motion.

Various clustering algorithms have been studied in the past with the purpose of finding a valid non-parameterized method.
UPMASK~\cite{krone2014upmask} was developed to only use photometry and positions,
and then it was adapted to the Gaia data~\cite{cantat2018gaia} based only on proper motion and parallax.
A similar approach has been carried out using DBSCAN and machine learning algorithms to discover new clusters.

@@

\begin{displayquote}
The non-field population does not occupy the entire workspace, but is spatially concentrated,
which makes it possible to distinguish two regions in the workspace:
the only field region (label `f`), dominated by star fields,
and the cluster + field region (label `c+f`), which includes both star fields and not star
fields~\cite{balaguer2020clusterix}.
\end{displayquote}

This fact implies defining three areas or regions with different radius.
The first `c+f` corresponds to the one in which the cluster members are
presumed to be contained together with other star fields that are not part of the cluster.
The second region is the broadest and assumes that it only contains stars
in an extended visual field without components of the cluster.
The third region is the intermediate one and is out of analysis (void area),
since it would correspond to a possible transition zone between the other two.

The right choice of these radii, even having a previous estimation for the `c+f` region,
highly affects the execution of the algorithm and, in general, requires a considerable wide field `f`.
There is no rule of thumb that defines relative proportions of these areas.

Finally, when an acceptable result is obtained,
what Clusterix provides is a probability field associated to the set of stars under study.
The recovered dataset contains all fields from Gaia database
and an extra column with the probability value of each object to belong to the open cluster.
This new dataset can be exported to TOPCAT for further optimization and final processing.
Figure~\ref{fig:clusterix_probability} shows an example of the probability distribution
in the proper motion space for NGC 2682.

@@

Finally, it is necessary to refer to the work developed on~\cite{castro2020hunting}.
It presents a method based on machine learning techniques to make a systematic search
for overdensities in the astrometric space of the galactic disk
and a later identification of OCs using photometric information, also from Gaia DR2.

The method includes two phases: the first one uses an unsupervised clustering algorithm, DBSCAN,
to search for overdensities \((l, b \pi, \mu_{\alpha} *, \mu_{\delta})\),
and then applies a deep learning Artificial Neural Network (ANN),
previously trained with magnitude diagrams,
to identify isochrone patterns within the detected overdensities and thus proceed to confirm them as OC.

@@

The result of this work is the recognition of 582 new open clusters
distributed along the galactic disk for a galactic declination \(b < 20^{\circ}\),
which has meant an increase of 45\% over previously ones.

Our aim in this work, linked to computational limitations compared to MareNostrum,
is not to perform a blind search for clusters, but to obtain a method, within the machine learning environment,
that allows the characterization of open clusters through unsupervised and non-parameterized procedures in a single step.
This method should not require further refinements with other tools.

@@

There are several clustering algorithms: \emph{K-Means}, \emph{Mean-Shift Clustering}, \emph{DBSCAN}~\cite{ester1996density} among other.
While each one behaves better according to the distribution of the objects to clusterize,
we have chosen K-Means by its simplicity and good results.

K-Means requires a single parameter, the number of clusters to build (\(N\)).
The algorithm looks for N center points which are vectors of the same dimension as the number of selected features.
K-Means starts by making an initial groups configuration and then,
it reassigns objects to other groups iteratively by minimizing the distance among points
inside the new group and maximizing the distance with the centers of the other groups.

@@

This algorithm works well and gives good results at first approach.
However, we need to set a large number of clusters to find the open cluster we are looking for.
This fact complicates the identification of the OC and many times the found cluster contains to many outliers.
For that reason, we searched for a K-Means refinement based on an artificial neural network.

The \emph{Unsupervised Deep Embedding for Clustering Analysis} model
or \emph{DEC}~\cite{xie2016unsupervised} takes K-Means as its starting point,
but then, it trains an autoencoder to reduce the feature space
and pass this transformed data through a Clustering Layer which refines the previous selection.

The mechanism of this model is explained in Section~\ref{sec:deep_embedding_clustering}
and its implementation is available in \verb|cdalvaro.ml.dec| package.

@@

All code has been developed with the Python programming language~\cite{Python3}.
Other auxiliary tools like \emph{Docker}~\cite{merkel2014docker},
\emph{PostgreSQL}~\cite{postgresql}, \emph{Jupyter Notebooks}~\cite{Kluyver2016jupyter},
and frameworks such as \emph{Astropy}~\cite{astropy:2013}~\cite{astropy:2018},
\emph{Scikit-Learn}~\cite{scikit-learn}, \emph{Seaborn}~\cite{michael_waskom_2017_883859},
\emph{SQLAlchemy}~\cite{sqlalchemy} and \emph{Keras}~\cite{chollet2015keras} have been used too.

@@

For the sake of simplicity, we are taking \emph{Melotte 22 (Messier 45)}~\cite{elsanhoury2019ppmxl} to illustrate the whole method.
This cluster is well studied and its stars are not too mixed with other stars that do not belong to the OC
although they are contained in the same observation field, so images look clear.

Figure~\ref{fig:raw_pm_melotte_22} shows \emph{proper motion in right ascension and declination}
for a sample of the downloaded dataset for Melotte 22.
At first sight, two main clusters can be distinguished, one of them centered nearly at (0, 0)
and the second one with center at (20, -45). This second cluster is the one we are looking for.

@@

Figure~\ref{fig:raw_parallax_hr_diagram_mellote_22} shows the parallax histogram and the H-R diagram for this cluster, respectively.

The figure on the left shows a resonance at \(\approx 7.3mas\) belonging the OC.
While the figure on the right would help us to look for isochrone curves and so,
to determine the age of these stars.

@@

Before being able to develop and test our clustering model,
the first step is to download the required data from the Gaia repository
and store it in a custom database for later access.

Due to the large amount of data available at Gaia, a complete download is not viable neither useful.
In order to reduce the size of the dataset to be downloaded,
the OpenClust catalogue~\cite{dias2002new} (Figure~\ref{fig:OpenClustComplete})
has been used to restrict the sky regions to be explored.

@@

As shown in Figure~\ref{fig:OpenClustSelection},
these constrains give us a smaller dataset but it is still a good representation of all clusters in the Milky Way
since they are equally distributed around the galaxy disk.

@@

The first one is a container built from a \verb|python:3.8| image that contains the \verb|cdalvaro| package
and the \verb|downloader.py| script.
This script is prepared to load the OpenClust catalogue, connect to the Gaia DR2 database,
download all stars contained inside each cluster (with a radius factor to increase the area to be downloaded)
and save the downloaded data inside our custom database hosted in the second container.

@@

These containers are orchestrated with a \verb|docker-compose.yml| file which automatically launches one instance of each container.
Figure~\ref{fig:services_diagram} shows a diagram describing the services architecture for this work.

@@

The second table is partitioned by region,
storing all stars related to one region inside an independent table for optimized access performance.
Partition tables are located inside \verb|gaiadr2| schema. See Figure~\ref{fig:customdb_diagram}.

@@

Table~\ref{tab:important_fields} shows a brief description of the fields used in this work.

@@

As mentioned before, we want our model to characterize open clusters by looking at their dynamic properties.
Also, we want to maintain as simple as possible our clustering model in order to save computing resources.

@@

Parallax is another important feature. It lets us know how far stars are from the Earth.
In addition, since all stars within an open cluster were born from the same dust cloud,
they must all have similar parallax.

@@

Figure~\ref{fig:features_melotte_22} shows a pairwise relationship among some features available
in the dataset for Melotte 22 data.
The diagonal of the grid shows the marginal distribution of the data in each column.
This plot can be useful to find correlations between pairs of variables or
to identify resonances in certain variables such as those in the parallax feature.

@@

\begin{itemize}
  \item \emph{Proper motion in right ascension} (corrected by parallax): \(\mu_{\alpha}\)
  \item \emph{Proper motion in declination} (corrected by parallax): \(\mu_{\delta}\)
  \item \emph{Parallax}: \( \varpi \)
  \item \emph{Proper motion modulus}: \(\left\| \vec{\mu} \right\|\)
\end{itemize}

This feature selection has been refined by iterating over the K-Means clustering process
with a fixed number of clusters and varying the feature selection in each case.
The final selection is the one with best \emph{silhouette score}~\cite{rousseeuw1987silhouettes}.

@@

Once we have found the set of features that best describes our problem,
we can begin searching for the OC within the selected region.

@@

As shown in Figure~\ref{fig:kmeans_comparisons_melotte_22},
larger values for the number of clusters allow us to isolate more accurately the resonance in parallax
at \(\approx 7.3mas\). However, we have the disadvantage that more groups are formed.

@@

K-Means does a good job making an initial clustering,
as shown in Figure~\ref{fig:kmeans_melotte_22}.
However, too many clusters arise from this characterization
and the OC is still polluted with stars that do not belong to it.
Moreover, we would like to reduce the amount of clusters too.

@@

If we take a look into the H-R diagram (Figure~\ref{fig:kmeans_hr_diagram_melotte_22}),
we can identify the isochrone curve for Melotte 22 cluster.
This curve has a good shape, but again, it contains outsider stars.
Therefore, we would like to find a better model that improves this initial characterization
by reducing the amount of clusters and also that removes outsiders from the OC.

@@

The model is composed by a \emph{deep autoencoder} and a \emph{clustering layer}.

The autoencoder is used to transform the input data into a latent space
using a non-linear mapping function \(f_{\theta} : X \rightarrow Z\).

Although, as explained in Section~\ref{sec:feature_selection},
the number of features we are managing is not too large,
this latent space helps us reduce the number of features
and avoids the \emph{``curse of dimensionality``}~\cite{bellman1961curse}.

The autoencoder is pretrained before fitting the model to generate predictions. Then,
the encoder layers of the autoencoder are used with the aim of transforming input data to
the latent space \(Z\). Once the data has been transformed, a K-Means clusterer is used in
order to make an initial clustering. K-Means cluster centers are used as the initial
weights for the clustering layer.

With that initial configuration,
the model iterates alternating between computing an auxiliary target distribution (Soft Assignment)
and minimizing the Kullback-Leiber (KL) divergence~\cite{kullback1951information} to it.
This unsupervised algorithm allows us to improve the clustering.

@@

In the soft assignment stage,
the \emph{Student's t-distribution} is used as a kernel to measure the similarity
between the embedded points and the cluster centroid.
While in the KL divergence minimization the algorithm iteratively refines clusters by learning
from their high confidence assignments with the help of an auxiliary target distribution.
The model is trained by matching the soft assignment to the target distribution.
The choice of this target distribution is crucial for DEC's performance.
In this work we have taken the target distribution from DEC's original paper~\cite{xie2016unsupervised},
which is defined in Equation~\ref{eq:student_tdistribution}.

Figure~\ref{fig:dec_model_setup} shows the layer setup of our DEC model.
It is simpler than the one tested on the original paper~\cite{xie2016unsupervised},
since the number of selected features in our work is smaller than in the original one.
Therefore, using the same configuration would result in a model so
powerful that would incur in overfitting issues unable to make right predictions.

@@

As shown in Figure~\ref{fig:dec_melotte_22},
the DEC model is able to move stars from one group to another,
removing some clusters and improving the OC characterization.
This is exactly what we are looking for.

@@

The H-R diagram has been improved as well,
and the isochrone curve of the open cluster is sharper than the one obtained with the K-Means algorithm
(Figure~\ref{fig:kmeans_hr_diagram_melotte_22}).

Figure~\ref{fig:hr_diagram_comparison} shows a comparison between the H-R diagram obtained
by DEC clustering and the one obtained using the Clusterix + TOPCAT method.

@@

Finally, we can refine this selection by filtering those stars which are bellow and above
the 0.10 and 0.90 quantiles for each group, respectively.
That way we remove the most doubtful values from the selection.
Figure~\ref{fig:melotte_22_filtered} shows the result after filtering DEC clustering selection.
